Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-04-06 inc/model_results/shapes/shapes.tex
move punctuation to after citation

For an electron pulse with a higher initial space-charge density, intra-pulse Coulomb effects may dominate the propagation dynamics.
\ref{fig:compare_shape}\subref{fig:compare_shape_N1e6}, shows the predicted time evolution dynamics of electron pulses with the same initial size and photoemission conditions as \ref{fig:compare_shape}\subref{fig:compare_shape_N1}, but with $ N = 10^{6} $ electrons --- the number required for single-shot diffraction pattern measurements in a DTEM.\cite{berger_dc_2009,armstrong_practical_2007}
The significant influence of space-charge effects for the higher initial peak pulse charge density is clearly evident.
In the spherical case, space-charge induced bunch broadening begins to overwhelm the initial difference between $\eta_{\smallT}$ and $\eta_{z}$, producing a more uniform three-dimensional `Coulomb explosion' of the electron pulse with the overall bunch broadening rate much greater than that for $ N = 10^{0} $ (\ref{fig:compare_shape}\subref{fig:compare_shape_N1}).
For $ \xi (0) > 1 $ (or $ \xi (0) < 1 $), the smaller initial transverse (longitudinal) bunch dimension experiences the greatest relative expansion in free-space propagation. This is to be expected, because the initial $\eta_{\alpha}$ are non-zero (but comparable in magnitude) while Coulomb effects will produce the greatest differential pulse broadening in the dimension with the largest internal space-charge field (i.e., the dimension with the largest electrostatic potential gradient).
% the following sentance is no longer true, still perhaps should include something like it
%In fact, both the $ N = 10^{6} $ oblate and prolate pulse examples shown in \ref{fig:compare_shape}\subref{fig:compare_shape_N1e6} transition through a spherical bunch shape ($ \xi \approx 1 $) after a time of flight of $\sim$0.5ns, at which point the peak pulse charge density has already been reduced by a factor of $\sim$100.
The initial impulse induced by the strong space-charge field then provides for continued linear expansion (in time), resulting in a linear pulse chirp in the dimension associated with the smallest initial HW1/eM bunch width\cite{michalik_analytic_2006,berger_dc_2009} $ \sqrt{ 2 \sigma_{\alpha} (0) } $.

The data presented in \ref{fig:compare_shape} using the AG model of Michalik and Sipe \cite{michalik_analytic_2006} are entirely consistent with prior works.\cite{reed_femtosecond_2006,siwick_ultrafast_2002}
In femtosecond UED experiments, significant efforts are made to offset space-charge induced temporal (longitudinal, $z$) pulse broadening by reducing the time of flight from the gun photocathode to the sample.\cite{siwick_ultrafast_2002,reed_evolution_2009}
Moreover, efforts are now underway to compensate for temporal electron pulse broadening using RF cavity pulse compression techniques.\cite{oudheusden_electron_2007}
In UEM, employing a standard retro-fitted electron microscope column driven by MHz repetition-rate sub-picosecond lasers,\cite{lobastov_four-dimensional_2005} deleterious space-charge effects have obliged operation in the $\sim$1 electron/pulse regime to maintain high spatial resolution.
Even for nanosecond DTEM (the extreme prolate pulse case, $\xi \gg 1 $), transverse space-charge induced beam broadening has necessitated the insertion of an additional magnetic electron lens in a microscope column to allow more efficient delivery of electrons to the specimen when $ N > 10^{6} $.\cite{lagrange_nanosecond_2008}
In this thesis, however, I will often restrict the discussion to ultrashort (ps and sub-ps) electron pulses which (i) allow for the temporal resolution of fundamental events occurring on ultrafast time scales in materials science, biology and chemistry \cite{king_ultrafast_2005} and (ii) can be easily generated in the oblate regime ($ \xi < 1 $) where transverse bunch broadening effects are minimized (\ref{fig:compare_shape})(enabling more efficient beam propagation) and RF cavities can be used to compensate and reverse longitudinal (temporal) pulse broadening.\cite{veisz_hybrid_2007}

>>>>

For an electron pulse with a higher initial space-charge density, intra-pulse Coulomb effects may dominate the propagation dynamics.
\ref{fig:compare_shape}\subref{fig:compare_shape_N1e6}, shows the predicted time evolution dynamics of electron pulses with the same initial size and photoemission conditions as \ref{fig:compare_shape}\subref{fig:compare_shape_N1}, but with $ N = 10^{6} $ electrons --- the number required for single-shot diffraction pattern measurements in a DTEM \cite{berger_dc_2009,armstrong_practical_2007}.
The significant influence of space-charge effects for the higher initial peak pulse charge density is clearly evident.
In the spherical case, space-charge induced bunch broadening begins to overwhelm the initial difference between $\eta_{\smallT}$ and $\eta_{z}$, producing a more uniform three-dimensional `Coulomb explosion' of the electron pulse with the overall bunch broadening rate much greater than that for $ N = 10^{0} $ (\ref{fig:compare_shape}\subref{fig:compare_shape_N1}).
For $ \xi (0) > 1 $ (or $ \xi (0) < 1 $), the smaller initial transverse (longitudinal) bunch dimension experiences the greatest relative expansion in free-space propagation. This is to be expected, because the initial $\eta_{\alpha}$ are non-zero (but comparable in magnitude) while Coulomb effects will produce the greatest differential pulse broadening in the dimension with the largest internal space-charge field (i.e., the dimension with the largest electrostatic potential gradient).
% the following sentance is no longer true, still perhaps should include something like it
%In fact, both the $ N = 10^{6} $ oblate and prolate pulse examples shown in \ref{fig:compare_shape}\subref{fig:compare_shape_N1e6} transition through a spherical bunch shape ($ \xi \approx 1 $) after a time of flight of $\sim$0.5ns, at which point the peak pulse charge density has already been reduced by a factor of $\sim$100.
The initial impulse induced by the strong space-charge field then provides for continued linear expansion (in time), resulting in a linear pulse chirp in the dimension associated with the smallest initial HW1/eM bunch width\cite{michalik_analytic_2006,berger_dc_2009} $ \sqrt{ 2 \sigma_{\alpha} (0) } $.

The data presented in \ref{fig:compare_shape} using the AG model of Michalik and Sipe \cite{michalik_analytic_2006} are entirely consistent with prior works \cite{reed_femtosecond_2006,siwick_ultrafast_2002}.
In femtosecond UED experiments, significant efforts are made to offset space-charge induced temporal (longitudinal, $z$) pulse broadening by reducing the time of flight from the gun photocathode to the sample \cite{siwick_ultrafast_2002,reed_evolution_2009}.
Moreover, efforts are now underway to compensate for temporal electron pulse broadening using RF cavity pulse compression techniques \cite{oudheusden_electron_2007}.
In UEM, employing a standard retro-fitted electron microscope column driven by MHz repetition-rate sub-picosecond lasers \cite{lobastov_four-dimensional_2005}, deleterious space-charge effects have obliged operation in the $\sim$1 electron/pulse regime to maintain high spatial resolution.
Even for nanosecond DTEM (the extreme prolate pulse case, $\xi \gg 1 $), transverse space-charge induced beam broadening has necessitated the insertion of an additional magnetic electron lens in a microscope column to allow more efficient delivery of electrons to the specimen when $ N > 10^{6} $ \cite{lagrange_nanosecond_2008}.
In this thesis, however, I will often restrict the discussion to ultrashort (ps and sub-ps) electron pulses which (i) allow for the temporal resolution of fundamental events occurring on ultrafast time scales in materials science, biology and chemistry \cite{king_ultrafast_2005} and (ii) can be easily generated in the oblate regime ($ \xi < 1 $) where transverse bunch broadening effects are minimized (\ref{fig:compare_shape})(enabling more efficient beam propagation) and RF cavities can be used to compensate and reverse longitudinal (temporal) pulse broadening \cite{veisz_hybrid_2007}.

### 2013-04-06 inc/ms_model/ms_model.tex
move punctuation to after citation

This AG formalism of Michalik and Sipe \cite{michalik_analytic_2006} employs a mean-field approximation in the evolution of the dynamical effect of the electron pulse's internal space-charge field.
This is of course not strictly accurate for a Gaussian charge distribution and the error leads to a distortion of the pulse shape.
The ``self-similar'' nature of the model, the fact that the overall shape of the pulse cannot change, prevents inclusion of these distortions.
Even so, the AG model of charge bunch dynamics has been benchmarked against particle tracking simulations for a number of electron pulse shapes,\cite{michalik_analytic_2006,michalik_evolution_2009} including the uniform ellipsoid which explicitly features a linear internal space-charge field.\cite{luiten_how_2004} 
%TODO add the error found in these benchmarks

>>>>

This AG formalism of Michalik and Sipe \cite{michalik_analytic_2006} employs a mean-field approximation in the evolution of the dynamical effect of the electron pulse's internal space-charge field.
This is of course not strictly accurate for a Gaussian charge distribution and the error leads to a distortion of the pulse shape.
The ``self-similar'' nature of the model, the fact that the overall shape of the pulse cannot change, prevents inclusion of these distortions.
Even so, the AG model of charge bunch dynamics has been benchmarked against particle tracking simulations for a number of electron pulse shapes \cite{michalik_analytic_2006,michalik_evolution_2009}, including the uniform ellipsoid which explicitly features a linear internal space-charge field \cite{luiten_how_2004}.
%TODO add the error found in these benchmarks

### 2013-04-06 inc/photocathode/setup/setup.tex
move punctuation to after citation

%Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED)\cite{sciaini_femtosecond_2011}, ultrafast transmission electron microscopy (UEM)\cite{king_ultrafast_2005}, and pulsed X-ray free electron lasers\cite{dowell_cathode_2010}.
The spatial quality of the sources for UEM and UED are often characterized in terms of their normalized transverse rms emittance\cite{dowell_quantum_2009,jensen_emittance_2010}; $\varepsilon_{\smallT} = \Delta p_x \Delta x / (m_0 c)$, where $\Delta p_x = \sqrt{ \langle p_x^2 \rangle }$ is the rms transverse momentum of the source and $\Delta x = \sqrt{ \langle x^2 \rangle }$  is its size (cylindrical symmetry assumed), and $m_0$ is the electron rest mass and $c$ is the speed of light in vacuum.
This quantity is conserved under free-space propagation and perfect lensing, and may increase due to lens aberrations \cite{oshea_reversible_1998}.
As I have discussed in Chapter \ref{chap:considerations}, a electron beam (or in the case of UEM, a pulse) with a lower transverse emittance will result in higher quality images, however since this quantity is conserved, the only chance to minimize this quantity occurs upon emission.
As a reduction in the source size $\Delta x$ is limited either by the attainable diffraction-limited focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field)\cite{valfells_effects_2002}, significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness\cite{berger_dc_2009}) may only be possible by decreasing the rms transverse momentum $\Delta p_x$ of the electron source --- an intrinsic property of the emission source\cite{dowell_quantum_2009,jensen_emittance_2010}.

@@

\ref{fig:transverse-measurement} depicts the experimental technique used to determine directly the transverse momentum distributions (and hence $\Delta p_x$) of the laser-driven electron sources.
Though many of the experimental components have been described in detail in Chapter \ref{chap:prototype}, for convenience I will briefly summarize them again here.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator\cite{berger_high-power_2008} delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate.
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
Both the green and UV laser pulses are focused with 30cm focal lenses onto the photocathode in a 20kV electron gun at an angle of incidence of $60(\pm5)^{\circ}$, which improves the coupling of the $p$-polarized laser radiation into the photocathode surface\cite{berger_dc_2009}.
The electron gun design is based on the that of Togawa et al.\cite{berger_dc_2009,togawa_ceb6_2007} and has an acceleration gap of 25($\pm$2)mm.
After acceleration in the DC photo-gun, the electrons pass through a set of electrostatic deflection plates to ensure that the electron beam passes, as close as possible, axially through the center of two, 6.35mm-bore, thin and round magnetic lenses with counter-propagating currents to minimize image rotation effects.
A YAG scintillator placed 27.5cm after the second magnetic lens detects the spatial electron beam profile, and its visible fluorescence is 1:1 imaged onto a CCD detector with 5.4$\mu$m pixels.

>>>>

%Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED) \cite{sciaini_femtosecond_2011}, ultrafast transmission electron microscopy (UEM)\cite{king_ultrafast_2005}, and pulsed X-ray free electron lasers\cite{dowell_cathode_2010}.
The spatial quality of the sources for UEM and UED are often characterized in terms of their normalized transverse rms emittance \cite{dowell_quantum_2009,jensen_emittance_2010}; $\varepsilon_{\smallT} = \Delta p_x \Delta x / (m_0 c)$, where $\Delta p_x = \sqrt{ \langle p_x^2 \rangle }$ is the rms transverse momentum of the source and $\Delta x = \sqrt{ \langle x^2 \rangle }$  is its size (cylindrical symmetry assumed), and $m_0$ is the electron rest mass and $c$ is the speed of light in vacuum.
This quantity is conserved under free-space propagation and perfect lensing, and may increase due to lens aberrations \cite{oshea_reversible_1998}.
As I have discussed in Chapter \ref{chap:considerations}, a electron beam (or in the case of UEM, a pulse) with a lower transverse emittance will result in higher quality images, however since this quantity is conserved, the only chance to minimize this quantity occurs upon emission.
As a reduction in the source size $\Delta x$ is limited either by the attainable diffraction-limited focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field)\cite{valfells_effects_2002}, significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness\cite{berger_dc_2009}) may only be possible by decreasing the rms transverse momentum $\Delta p_x$ of the electron source --- an intrinsic property of the emission source\cite{dowell_quantum_2009,jensen_emittance_2010}.

@@

\ref{fig:transverse-measurement} depicts the experimental technique used to determine directly the transverse momentum distributions (and hence $\Delta p_x$) of the laser-driven electron sources.
Though many of the experimental components have been described in detail in Chapter \ref{chap:prototype}, for convenience I will briefly summarize them again here.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator\cite{berger_high-power_2008} delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate.
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
Both the green and UV laser pulses are focused with 30cm focal lenses onto the photocathode in a 20kV electron gun at an angle of incidence of $60(\pm5)^{\circ}$, which improves the coupling of the $p$-polarized laser radiation into the photocathode surface\cite{berger_dc_2009}.
The electron gun design is based on the that of Togawa \textit{et al.} \cite{berger_dc_2009,togawa_ceb6_2007} and has an acceleration gap of 25($\pm$2)mm.
After acceleration in the DC photo-gun, the electrons pass through a set of electrostatic deflection plates to ensure that the electron beam passes, as close as possible, axially through the center of two, 6.35mm-bore, thin and round magnetic lenses with counter-propagating currents to minimize image rotation effects.
A YAG scintillator placed 27.5cm after the second magnetic lens detects the spatial electron beam profile, and its visible fluorescence is 1:1 imaged onto a CCD detector with 5.4$\mu$m pixels.
